A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-01 tex/MercuryNumerical.tex
added full example code and added coordinate system to figure
plotted turn variation

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories observed.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{dt}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

@@

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
	\todo[inline]{@CK: figure: coordinate system}
	\caption{\label{fig:sun_merc}%
		Sun mercury system with relevant vectors.
		Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 10
\end{lstlisting}
Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $dt$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

@@

\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\caption{\label{fig:MercuryOrbit}%
		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
		All computations where executed with $dt = 2 * v_{M}(0) / c_a / 10$.
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=.8\textwidth]{figs/AlphaAngle.png}
	\todo[inline]{@CK: pdf figure + confirm @IH: distribute data}
	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion.}
\end{figure}

To extract the perihelion motion at $\alpha = 1$, we use the fact that $\delta \Theta$ is zero for $\alpha =0$ --- if one could completely eliminate numerical errors.
Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.1^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=1) = \frac{10.1^{\degree}}{10^6} = 0.036''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
Therefore, the perihelion motion per 100 earth years is $0.036''\cdot 415\approx 15''$.
To test whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for mercury as outlined above.

@@

Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $dt$ and $\alpha$.
As expected, the correlation between the number of turns and $\delta\Theta$ is linear and the slope depends solely on $\alpha$.
However, there is an offset between the lines for different $dt$.
The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
This does not generate a problem for the procedure described in the previous section since the starting position is not considered. Still, it is helpful to be aware of this fact.
One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.

\begin{figure}[htb]
	\centering
	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
	\todo[inline]{@CK: Could you redo this figure with some bigger time steps? Could we call the precession growth $\Theta(\alpha)$ instead of $\delta\Theta$? I think this would fit better with our notation so far...}
	\caption{\label{fcc3}Precession growth of perihelion for $\delta\Theta$ in degrees over number of turns for different time steps (color).  One should pay special attention to the oscillations within the growth for relatively large time steps.}
\end{figure}

To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / c_a \cdot 5$ is shown.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $dt$.
Only in the limit $dt \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/num-err.png}
	\caption{\label{fig: LargeDT}%
		Deviation of the trajectory from precise numeric result (black line; $dt=2 v_M(0) / a_M / 20$) for large time steps (red line; $dt=2 v_M(0) / a_M \times 2$).
		Both orbits are plotted for $T = 10 T_M$.
	}
\end{figure}

@@

However in practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously.
By including \texttt{import time}
\begin{lstlisting}
start_time = time.time()
main()
print("--- %s seconds ---" % (time.time() - start_time))
\end{lstlisting}
the students can measure the time needed by their program.
By varying $dt$ they can validate, that there is indeed approximately an anti-proportional dependence.
(Note: This only works if the time in the loop is increased by $dt$, so $t=t+dt$.)
\item \textbf{Verlet integration}

Of course by optimizing the code, an improvement in accuracy can be achieved without changing $dt$.
The simplest way to demonstrate this might be given by the implementation of the Verlet\cite{}\todo{@IH: cite Verlet} integration instead of using the simple Euler method.
\end{itemize}

>>>>

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories observed.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

@@

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun mercury system with relevant vectors.
		Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 10
\end{lstlisting}
Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

@@

\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\caption{\label{fig:MercuryOrbit}%
		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
		All computations where executed with $\Delta t = 2 v_{M}(0) / c_a / 20$.
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

\begin{figure}[htb]
	\centering
	\includegraphics{figs/alpha-angle.pdf}
	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$.}
\end{figure}

To extract the perihelion motion at $\alpha = 1$, we use the fact that $\delta \Theta$ is zero for $\alpha =0$ --- if one could completely eliminate numerical errors.
Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.0^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=1) = \frac{10.0^{\degree}}{10^6} = 0.036''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
Therefore, the perihelion motion per 100 earth years is $0.036''\cdot 415\approx 15''$.
To test whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for mercury as outlined above.

@@

Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $\Delta t$ and $\alpha$.
As expected, the correlation between the number of turns and $\delta\Theta$ is linear and the slope depends solely on $\alpha$.
However, there is an offset between the lines for different $\Delta t$.
The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
This does not generate a problem for the procedure described in the previous section since the starting position is not considered. Still, it is helpful to be aware of this fact.
One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\todo[inline]{@all: where is this periodic oscillation coming from?}
	\caption{\label{fcc3}Precession growth of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
	One should pay special attention to the oscillations within the growth for relatively large time steps.
}
\end{figure}

To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
In figure \ref{fig: LargeDT} the mercury trajectory for $\Delta t=2 * v_M(0) / c_a \cdot 5$ is shown.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $\Delta t$.
Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/num-err.png}
	\caption{\label{fig: LargeDT}%
		Deviation of the trajectory from precise numeric result (black line; $\Delta t=2 v_M(0) / a_M / 20$) for large time steps (red line; $\Delta t=2 v_M(0) / a_M \times 2$).
		Both orbits are plotted for $T = 10 T_M$.
	}
\end{figure}

@@

However in practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously.
By including \texttt{import time}
\begin{lstlisting}
start_time = time.time()
main()
print("--- %s seconds ---" % (time.time() - start_time))
\end{lstlisting}
the students can measure the time needed by their program.
By varying $\Delta t$ they can validate, that there is indeed approximately an anti-proportional dependence.
(Note: This only works if the time in the loop is increased by $\Delta t$, so $t=t+\Delta t$.)
\item \textbf{Verlet integration}

Of course by optimizing the code, an improvement in accuracy can be achieved without changing $\Delta t$.
The simplest way to demonstrate this might be given by the implementation of the Verlet\cite{}\todo{@IH: cite Verlet} integration instead of using the simple Euler method.
\end{itemize}
